Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome

### 2017-12-11 docs/manuscript.tex

Pleiotropy is the phenomenon whereby a single genetic variant influences
multiple phenotypes. The mechanisms behind it can arise through many
different mechanisms, for example a single variant influences multiple
genes; a single gene generates multiple gene products; a single gene
product has multiple cellular functions; or a single cellular function
influences multiple phenotypes (23). How we classify the pleiotropic
mechanism depends on the application, and in this paper we discuss
pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy
has also been known as mediated pleiotropy (24), because the genetic
variant influences one trait, which in turn influences a second trait.
It has also been termed, in some contexts, Type II pleiotropy (16) and
secondary pleiotropy (25). It is this form of pleiotropy that MR assumes
to make causal inference between two traits.

@@

The four nested models (IVW fixed effects, IVW random effects, Egger
fixed effects, Egger random effects) plus the three Rucker estimates
(point estimate, mean of the jackknife, median of the jackknife) provide
seven mean based estimators.

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (5,29). This has the
advantage that only half the instruments need to be valid, and the
estimate will remain unbiased. The weighted median estimate develops the
approach further to allow stronger instruments to contribute more
towards the estimate can be obtained by weighting the contribution of
each instrument by the inverse of its variance. The penalised weighted
median estimator introduces a further weight to the instruments,
penalising any instrument that contributes substantially towards the
heterogeneity statistic. Together, this provides three median based
estimators. Other estimation strategies not considered here, such as
LASSO regression, have also been developed for the situation where at
least half of the instruments are valid (30).

\textbf{Mode-based methods:} The mode based estimator clusters the
instruments into groups based on similarity of causal effects, and
returns the final causal effect estimate based on the cluster that has
the largest number of instruments (7). There are four implementations of
this method: the simple and the weighted mode, each weighted with or
without the assumption of no measurement error in the exposure estimates
(NOME). The simple mode is the unweighted mode of the empirical density
function of causal estimates, whereas the weighted mode is weighted by
the inverse variance of the outcome effect. The bandwidth parameter was
set to 1 by default.

@@

With genome-wide association studies growing ever larger, the
statistical power to detect significant associations that may be
influencing the trait downstream of many other pathways increases. For
example, if a SNP \(g_{A}\) influences trait \(A\), and trait \(A\)
influences trait \(B\), then a sufficiently powered GWAS will identify
the \(g_{A}\) as being significant for trait \(B\) (Figure 1a). Using
\(g_{A}\) as an instrument to test the causal effect of \(A\) on \(B\)
is perfectly valid. But the (incorrectly hypothesised) MR analysis of
trait \(B\) on trait \(A\) could erroneously result in the apparent
causal effect of \(B\) on \(A\). If \(g_{A}\) is only one of many known
instruments for \(B\), amongst which some are valid, it is to the
advantage of the researcher to exclude \(g_{A}\) from the analysis.

@@

For specific hypotheses we strongly recommend that estimates from all
sensitivity analyses are scrutinised and reported. The graph can be
queried directly using the cypher language at
\url{http://eve-neo4j.mrbase.org} or through a basic web interface at
\url{http://eve.mrbase.org}.

@@

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to predict the most appropriate of the 28 MR
strategies and instrument selection strategies to apply to a particular
summary-set based on its characteristics (Figure 2).

@@

As an example of the database we performed a look up of associations
between all traits and LDL cholesterol (as measured in the GLGC
consortium (41)), where the MoE result had a false discovery rate of
0.05 using the MoE approach. This returned 287 putative associations,
amongst which 111 involved LDL cholesterol influencing other traits and
176 involved other traits influencing LDL cholesterol. A large
proportion of these traits were metabolites, which are dominated by
lipid fractions. Filtering to exclude the metabolomic studies ((38,39))
returned 27 associations with 23 traits (Figure 4a). There was a strong
association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52,
\(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
after Steiger filtering, indicating the presence of horizontal
pleiotropy being present amongst the instruments.

We also performed a look up of associations that causally influence
years of schooling (42), or that years of schooling itself causally
influences. Using a false-discovery rate of 0.05, 45 traits were
returned as having some direction of causality with years of schooling
(Figure 4b). Several of these putative relationships appear to be
plausible, for example the influence of years of schooling on lower risk
of Alzheimer's disease, which has recently been reported (43). However,
it is clear that there are serious limitations in relying on MR as a
panacea for causal inference. For example, bi-directional causal
relationships with college completion exist because of the similarity in
trait definitions; and there are several traits (e.g.~childhood
intelligence and birth weight) which appear to be causally downstream of
years of schooling despite this being temporally impossible. Though
explanations could be conjured for this association, for example
parents' schooling influences childhood intelligence, or the genetic
instruments are shared across the two traits due to a third unmeasured
trait, an intuitive interpretation of causality cannot be applied
because the genetic instruments used proxy the liability to achieve some
level of educational attainment, and not necessarily the experience of
having attained a particular level of edication.

The facility to perform these scans has been automated in the MR of
`everything vs everything' (MR-EvE) web application,
\url{http://eve.mrbase.org}.

@@

Motivated by a recent flurry of activity in MR methodology development,
we have devised a machine learning approach, MR-MoE, that seeks to
predict the performance of each MR method, selecting the one which is
most likely to maximise power whilst minimising FDR for a specific
summary-set. The mixture of experts approach that we present here is
trained using random forest decision trees applied to extensive and
diverse simulations, and we demonstrate that it makes substantial
performance improvements over any other single method in the presence or
absence of extensive horizontal pleiotropy. As such, our method
contributes to the field of meta-learning but also to the rapidly
growing field of automatic model selection and hyper-parameter
optimization in machine learning (44). We applied the method to the
curated set of GWAS summary data present in MR-Base (12), generating
several million MR estimates. MR-MoE selected a method that indicated
some pattern of horizontal pleiotropy for almost 90\% of cases,
reinforcing the notion that it is the rule rather than the exception.

The trend of increasing GWAS sample sizes continues, but while the
opportunity that this affords MR to be furnished with more instruments
is typically welcomed, here we have shown that it is invalid
instruments, exhibiting horizontal pleiotropy or reverse causation, that
are more likely to be identified than valid ones. Strategising how MR is
to be used in practice, therefore, must consider horizontal pleiotropy
to be the rule rather than the exception. Instrumental variables in MR
are typically chosen blind, with GWAS significance being the only
criterion. We illustrated how confounders and reverse causal
associations can easily lead to invalid instrument selection in MR and
that most randomly generated patterns of horizontal pleiotropy cannot be
adequately accounted for by any single MR method.

@@

A large proportion of the associations were estimated using only a
single instrumental variable, a circumstance in which cause, reverse
cause or horizontal pleiotropy cannot be immediately distinguished.
Though methods are emerging to attempt to delineate between models of
reverse cause, pleiotropy and multiple causal variants in the same
region (45,46), separating vertical from horizontal pleiotropy with a
single instrument is not yet possible in the two-sample MR framework
though the use of mediation-based approaches may be informative in the
one-sample setting (8,47).

Other problems, in addition to horizontal pleiotropy, can also manifest.
Frailty effects (48) could induce associations for late-onset traits,
and often detailed modelling for how this can influence MR is warranted
(43,49). Genetic variants could strongly relate to several phenotypes,
for example SNPs that influence LDL cholesterol are also likely to
influence other lipid fractions, making it difficult to ascertain which
amongst them is the true causal exposure (50). We see this issue
manifesting in Figure 4, where there appear to be causal relationships
between different lipid fractions. The measured feature for which
genetic associations are known could relate in complicated ways to the
biological entities that is truly causal (2), for example higher
circulating levels of the natural antioxidant extracellular superoxide
dismutase (EC-SOD) appears to causally relate to higher risk of coronary
heart disease (51) because the phenotypic measure relates inversely to
the EC-SOD levels in arterial walls, hence \emph{in situ} activity is
lower (2). Dynastic effects can also confound MR estimates, arising when
the exposure trait in a previous generation influences the environment
of the current generation. Here, the SNPs being used as instruments for
the exposure will be associated with a potential confounder in the
analysis (52). The associations depicted in Figure 4 involving
educational attainment may be susceptible to bias arising from this
mechanism, and protecting against it will likely require alternative
study designs, for example by using sibling pairs.

@@

The field of MR is evolving. The advent of GWAS databases (10,53) and
automated 2SMR (12) has trivialised the analytical aspect of
investigating specific causal hypotheses. Despite the limitations
described above, the construction of a causal graph of `everything
versus everything' does have appeal. First, though causality is not
guaranteed by MR, it can still be highly informative for supporting or
negating specific hypotheses. Indeed, a well-powered MR that negates a
hypothesis is likely less fallible to the pitfalls described above than
using MR to confirm a positive association (54). Second, it paves the
way for new approaches to search for novel putative associations and
improve reliability in single analyses. The potential to exploit GWAS
summary data within the properties of graph databases to aid with both
of these endeavours could be transformative in biological research.

@@

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2f(1-f)\), where \(\beta\) is the log odds ratio and
\(f\) is the allele frequency of the SNP in the population (55). \(f\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\(f_{cc}\), and number of cases \(N_1\) and controls \(N_0\). A Z-score
to test the hypothesis that \(cor(g, x)\) and \(cor(g, y)\) are
significantly different can be generated using the Steiger test for
correlated correlations.

@@

Having generated random forest decision trees for each of the 28
strategies using 133,000 of the simulations, we then applied them to the
remaining 67,000 summary-sets to predict which method would have the
highest performance for each of the remaining summary-sets. Finally we
compare the performance of the method selected by the MoE against all
remaining strategies. The default settings for the \texttt{randomForest}
package in R (56) were used to train the models. MR-MoE 1.0 is
implemented in the TwoSampleMR R package available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
(12).

@@

\hypertarget{ref-Hartwig2017}{}
7. Hartwig FP, Davey Smith G, Bowden J. Robust Inference In Two-Sample
Mendelian Randomisation Via The Zero Modal Pleiotropy Assumption.
bioRxiv {[}Internet{]}. 2017; Available from:
\url{http://biorxiv.org/content/early/2017/04/10/126102}

@@

\hypertarget{ref-Hu2016}{}
23. Hu JX, Thomas CE, Brunak S. Network biology concepts in complex
disease comorbidities. Nature reviews Genetics {[}Internet{]}.
2016;17(10):615--29. Available from:
\href{http://www.nature.com/doifinder/10.1038/nrg.2016.87\%7B/\%\%7D5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}{http://www.nature.com/doifinder/10.1038/nrg.2016.87\{\textbackslash{}\%\}5Cnhttp://www.ncbi.nlm.nih.gov/pubmed/27498692}

@@

\hypertarget{ref-Paaby2013}{}
26. Paaby AB, Rockman MV. The many faces of pleiotropy {[}Internet{]}.
Vol. 29. 2013. pp. 66--73. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/23140989\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540\%20http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}{http://www.ncbi.nlm.nih.gov/pubmed/23140989 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540 http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}

@@

\hypertarget{ref-kotthoff2016auto}{}
44. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
2.0: Automatic model selection and hyperparameter optimization in WEKA.
Journal of Machine Learning Research. 2016;17:1--5.

\hypertarget{ref-Zhu2016}{}
45. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
Integration of summary data from GWAS and eQTL studies predicts complex
trait gene targets. Nature Genetics {[}Internet{]}. 2016
Mar;48(5):481--7. Available from:
\url{http://www.nature.com/doifinder/10.1038/ng.3538}

\hypertarget{ref-Richardson2017}{}
46. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
CL, et al. Causal epigenome-wide association study identifies CpG sites
that influence cardiovascular disease risk. bioRxiv {[}Internet{]}.
2017; Available from:
\url{http://biorxiv.org/content/early/2017/04/29/132019}

\hypertarget{ref-Wang2017}{}
47. Wang L, Michoel T. Efficient And Accurate Causal Inference With
Hidden Confounders From Genome-Transcriptome Variation Data. bioRxiv
{[}Internet{]}. 2017; Available from:
\url{http://www.biorxiv.org/content/early/2017/04/19/128496}

\hypertarget{ref-Aalen2015}{}
48. Aalen OO, Valberg M, Grotmol T, Tretli S. Understanding variation in
disease risk: the elusive concept of frailty. International journal of
epidemiology {[}Internet{]}. 2015 Aug;44(4):1408--21. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/25501685\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4588855}{http://www.ncbi.nlm.nih.gov/pubmed/25501685 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4588855}

\hypertarget{ref-Noyce2017}{}
49. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
E, et al. Estimating the causal influence of body mass index on risk of
Parkinson disease: A Mendelian randomisation study. Brayne C, editor.
PLOS Medicine {[}Internet{]}. 2017 Jun;14(6):e1002314. Available from:
\url{http://dx.plos.org/10.1371/journal.pmed.1002314}

\hypertarget{ref-Burgess2014a}{}
50. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
multivariable Mendelian randomization to disentangle the causal effects
of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
Available from:
\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}

\hypertarget{ref-Juul2003}{}
51. Juul K, Tybjaerg-Hansen A, Marklund S, Heegaard NHH, Steffensen R,
Sillesen H, et al. Genetically Reduced Antioxidative Protection and
Increased Ischemic Heart Disease Risk: The Copenhagen City Heart Study.
Circulation {[}Internet{]}. 2003 Dec;109(1):59--65. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/14662715\%20http://circ.ahajournals.org/cgi/doi/10.1161/01.CIR.0000105720.28086.6C}{http://www.ncbi.nlm.nih.gov/pubmed/14662715 http://circ.ahajournals.org/cgi/doi/10.1161/01.CIR.0000105720.28086.6C}

\hypertarget{ref-VonHinke2016}{}
52. Hinke S von, Davey Smith G, Lawlor D, Propper C, Windmeijer F.
Genetic markers as instrumental variables. J Health Econ.
2016;45:131--48.

\hypertarget{ref-Staley2016}{}
53. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
al. PhenoScanner: a database of human genotype--phenotype associations.
Bioinformatics {[}Internet{]}. 2016 Oct;32(20):3207--9. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/27318201\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5048068\%20https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btw373}{http://www.ncbi.nlm.nih.gov/pubmed/27318201 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5048068 https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btw373}

\hypertarget{ref-Vanderweele2014}{}
54. Vanderweele TJ, Tchetgen Tchetgen EJ, Cornelis M, Kraft P.
Methodological challenges in mendelian randomization. Epidemiology
(Cambridge, Mass) {[}Internet{]}. 2014 May;25(3):427--35. Available
from: \url{http://www.ncbi.nlm.nih.gov/pubmed/24681576}

\hypertarget{ref-Lee2013c}{}
55. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
association studies: quantification of power and genomic prediction
accuracy. PLoS One. 2013;8(8):e71494.

>>>>

Pleiotropy is the phenomenon whereby a single genetic variant influences
multiple phenotypes. The mechanisms behind it can arise through many
different mechanisms, for example a single variant influences multiple
genes; a single gene generates multiple gene products; a single gene
product has multiple cellular functions; or a single cellular function
influences multiple phenotypes (23). How we classify the pleiotropic
mechanism depends on question and context, and in this paper we discuss
pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy
has also been known as mediated pleiotropy (24), because the genetic
variant influences one trait, which in turn influences a second trait.
It has also been termed, in some contexts, Type II pleiotropy (16) and
secondary pleiotropy (25). It is this form of pleiotropy that MR assumes
to make causal inference between two traits.

@@

The four nested models (IVW fixed effects, IVW random effects, Egger
fixed effects, Egger random effects) plus the three Rucker estimates
(point estimate, mean of the jackknife, median of the jackknife) provide
seven mean-based estimators.

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (5,29). This has the
advantage that only half the instruments need to be valid, and the
estimate will remain unbiased. The weighted median estimate develops the
approach further to allow stronger instruments to contribute more
towards the estimate can be obtained by weighting the contribution of
each instrument by the inverse of its variance. The penalised weighted
median estimator introduces a further weight to the instruments,
penalising any instrument that contributes substantially towards the
heterogeneity statistic. Together, this provides three median-based
estimators. Other estimation strategies not considered here, such as
LASSO regression, have also been developed for the situation where at
least half of the instruments are valid (30).

\textbf{Mode-based methods:} The mode-based estimator clusters the
instruments into groups based on similarity of causal effects, and
returns the final causal effect estimate based on the cluster that has
the largest number of instruments (7). There are four implementations of
this method: the simple and the weighted mode, each weighted with or
without the assumption of no measurement error in the exposure estimates
(NOME). The simple mode is the unweighted mode of the empirical density
function of causal estimates, whereas the weighted mode is weighted by
the inverse variance of the outcome effect. The bandwidth parameter was
set to 1 by default.

@@

With genome-wide association studies growing ever larger, the
statistical power to detect significant associations that may be
influencing the trait downstream of many other pathways increases. For
example, if a SNP \(g_{A}\) influences trait \(A\), and trait \(A\)
influences trait \(B\), then a sufficiently powered GWAS will identify
the \(g_{A}\) as being significant for trait \(B\) (Figure 1a). Using
\(g_{A}\) as an instrument to test the causal effect of \(A\) on \(B\)
is perfectly valid in some cases. But the (incorrectly hypothesised) MR
analysis of trait \(B\) on trait \(A\), for example, could erroneously
result in the apparent causal effect of \(B\) on \(A\). It is to the
advantage of the researcher to exclude \(g_{A}\) from the analysis from
the analysis of \(B\) against \(A\).

@@

The graph can be queried directly using the cypher language at
\url{http://eve-neo4j.mrbase.org} or through a basic web interface at
\url{http://eve.mrbase.org}. For specific hypotheses we strongly
recommend that estimates from all sensitivity analyses are scrutinised
and reported.

@@

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to predict the most appropriate of the 14 MR methods
and two instrument selection strategies (giving a total of 28 MR
strategies) to apply to a particular summary-set based on its
characteristics (Figure 2).

@@

We performed a look up of associations between all traits and LDL
cholesterol (as measured in the GLGC consortium (41)), where the result
had a false discovery rate of 0.05 using the MoE approach. This returned
287 putative associations, amongst which 111 involved LDL cholesterol
influencing other traits and 176 involved other traits influencing LDL
cholesterol. A large proportion of these traits were metabolites, which
are dominated by lipid fractions. Filtering to exclude the metabolomic
studies ((38,39)) returned 27 associations with 23 traits (Figure 4a).
There was a strong association with coronary heart disease (0.41 log(OR)
per SD, 0.31-0.52, \(p=2.7\times10^{-12}\)). Here the IVW random effects
method was chosen after Steiger filtering, indicating the presence of
horizontal pleiotropy being present amongst the instruments.

We also performed a look up of associations that causally influence
years of schooling (42), or that years of schooling itself causally
influences. Using a false-discovery rate of 0.05, 45 traits were
returned as having some direction of causality with years of schooling
(Figure 4b). Several of these putative relationships appear to be
plausible, for example the influence of years of schooling on lower risk
of Alzheimer's disease, which has recently been reported (43). There is
also a protective effect of more years of schooling on lower risk of
myocardial infarction, which has been supported by detailed follow up
analyses (44).

However, it is clear that there are limitations in relying on MR as a
panacea for causal inference. For example, bi-directional causal
relationships with college completion exist because of the similarity in
trait definitions; and there are several traits (e.g.~childhood
intelligence and birth weight) which appear to be causally downstream of
years of schooling despite this being temporally impossible. Though
explanations could be conjured for this association, for example
parents' schooling influences childhood intelligence, or the genetic
instruments are shared across the two traits due to a third unmeasured
trait, an intuitive interpretation of causality cannot be applied
because the genetic instruments used proxy the liability to achieve some
level of educational attainment, and not necessarily the experience of
having attained a particular level of edication.

The facility to perform these look ups has been automated in the MR of
`everything vs everything' (MR-EvE) web application,
\url{http://eve.mrbase.org}.

@@

Motivated by a recent flurry of activity in MR methodology development,
we have devised a machine learning approach, MR-MoE, that seeks to
predict the performance of each MR method, selecting the one which is
most likely to maximise power whilst minimising FDR for a specific
summary-set. The mixture of experts approach that we present here is
trained using random forest decision trees applied to extensive and
diverse simulations, and we demonstrate that it makes substantial
performance improvements over any other single method in the presence or
absence of extensive horizontal pleiotropy. As such, our method
contributes to the field of meta-learning but also to the rapidly
growing field of automatic model selection and hyper-parameter
optimization in machine learning (45). We applied the method to the
curated set of GWAS summary data present in MR-Base (12), generating
several million MR estimates. MR-MoE selected a method that indicated
some pattern of horizontal pleiotropy for almost 90\% of cases,
reinforcing the notion that it is the rule rather than the exception.

The trend of increasing GWAS sample sizes continues, but while the
opportunity that this affords MR to be furnished with more instruments
is typically welcomed, here we have demonstrated that it is invalid
instruments, exhibiting horizontal pleiotropy or reverse causation, that
may be more likely to be identified than valid ones. Strategising how MR
is to be used in practice, therefore, must consider some presence of
horizontal pleiotropy to be the rule rather than the exception.
Instrumental variables in MR are typically chosen blind, with GWAS
significance being the only criterion. We illustrated how confounders
and reverse causal associations can easily lead to invalid instrument
selection in MR and that most randomly generated patterns of horizontal
pleiotropy cannot be adequately accounted for by any single MR method.

@@

A large proportion of the associations were estimated using only a
single instrumental variable, a circumstance in which cause, reverse
cause or horizontal pleiotropy cannot be immediately distinguished.
Though methods are emerging to attempt to delineate between models of
reverse cause, pleiotropy and multiple causal variants in the same
region (46,47), separating vertical from horizontal pleiotropy with a
single instrument is not yet possible in the two-sample MR framework
though the use of mediation-based approaches may be informative in the
one-sample setting (8,48).

Other problems, in addition to horizontal pleiotropy, can also manifest.
Frailty effects (49) could induce associations for late-onset traits,
and often detailed modelling for how this can influence MR is warranted
(43,50). Genetic variants could strongly relate to several phenotypes,
for example SNPs that influence LDL cholesterol are also likely to
influence other lipid fractions, making it difficult to ascertain which
amongst them is the true causal exposure (51). We see this issue
manifesting in Figure 4, where there appear to be causal relationships
between different lipid fractions. The measured feature for which
genetic associations are known could relate in complicated ways to the
biological entities that is truly causal (2), for example higher
circulating levels of the natural antioxidant extracellular superoxide
dismutase (EC-SOD) appears to causally relate to higher risk of coronary
heart disease (52) because the phenotypic measure relates inversely to
the EC-SOD levels in arterial walls, hence \emph{in situ} activity is
lower (2). Dynastic effects can also confound MR estimates, arising when
the exposure trait in a previous generation influences the environment
of the current generation. Here, the SNPs being used as instruments for
the exposure will be associated with a potential confounder in the
analysis (53). The associations depicted in Figure 4 involving
educational attainment may be susceptible to bias arising from this
mechanism, and protecting against it will likely require alternative
study designs, for example by using sibling pairs.

@@

The field of MR is evolving. The advent of GWAS databases (10,54) and
automated 2SMR (12) has trivialised the analytical aspect of
investigating specific causal hypotheses. Despite the limitations
described above, the construction of a causal graph of `everything
versus everything' does have appeal. First, though causality is not
guaranteed by MR, it can still be highly informative for supporting or
negating specific hypotheses. Indeed, a well-powered MR that negates a
hypothesis is likely less fallible to the pitfalls described above than
using MR to confirm a positive association (55). Second, it paves the
way for new approaches to search for novel putative associations and
improve reliability in single analyses. The potential to exploit GWAS
summary data within the properties of graph databases to aid with both
of these endeavours could be transformative in biological research.

@@

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2f(1-f)\), where \(\beta\) is the log odds ratio and
\(f\) is the allele frequency of the SNP in the population (56). \(f\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\(f_{cc}\), and number of cases \(N_1\) and controls \(N_0\). A Z-score
to test the hypothesis that \(cor(g, x)\) and \(cor(g, y)\) are
significantly different can be generated using the Steiger test for
correlated correlations.

@@

Having generated random forest decision trees for each of the 28
strategies using 133,000 of the simulations, we then applied them to the
remaining 67,000 summary-sets to predict which method would have the
highest performance for each of the remaining summary-sets. Finally we
compare the performance of the method selected by the MoE against all
remaining strategies. The default settings for the \texttt{randomForest}
package in R (57) were used to train the models. MR-MoE 1.0 is
implemented in the TwoSampleMR R package available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
(12).

@@

\hypertarget{ref-Hartwig2017a}{}
7. Hartwig FP, Smith GD, Bowden J. Robust inference in summary data
Mendelian randomization via the zero modal pleiotropy assumption.
International Journal of Epidemiology {[}Internet{]}. 2017;dyx102:1--14.
Available from:
\href{https://oup.silverchair-cdn.com/oup/backfile/Content\%7B/_\%7Dpublic/Journal/ije/PAP/10.1093\%7B/_\%7Dije\%7B/_\%7Ddyx102/2/dyx102.pdf?Expires=1502303601\%7B/\&\%7DSignature=FC72PSJXMhTtysiYRAtlBTDUthqtIDgOIbjxvoO\%7B~\%7Dn0oiMIEOdXKEpXu7Wxq0EKKPev57T9LPxLzh81sX7\%7B~\%7DYzHu0s6Fd41zYJUfdISx8ev65rchmVp4AMcM}{https://oup.silverchair-cdn.com/oup/backfile/Content\{\textbackslash{}\_\}public/Journal/ije/PAP/10.1093\{\textbackslash{}\_\}ije\{\textbackslash{}\_\}dyx102/2/dyx102.pdf?Expires=1502303601\{\textbackslash{}\&\}Signature=FC72PSJXMhTtysiYRAtlBTDUthqtIDgOIbjxvoO\{\textasciitilde{}\}n0oiMIEOdXKEpXu7Wxq0EKKPev57T9LPxLzh81sX7\{\textasciitilde{}\}YzHu0s6Fd41zYJUfdISx8ev65rchmVp4AMcM}

@@

\hypertarget{ref-Hu2016}{}
23. Hu JX, Thomas CE, Brunak S. Network biology concepts in complex
disease comorbidities. Nat Rev Genet {[}Internet{]}. 2016
Oct;17(10):615--29. Available from:
\href{http://dx.doi.org/10.1038/nrg.2016.87\%20http://10.0.4.14/nrg.2016.87}{http://dx.doi.org/10.1038/nrg.2016.87 http://10.0.4.14/nrg.2016.87}

@@

\hypertarget{ref-Paaby2013}{}
26. Paaby AB, Rockman MV. The many faces of pleiotropy. Trends in
Genetics {[}Internet{]}. 2013 Feb;29(2):66--73. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/23140989\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540\%20http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}{http://www.ncbi.nlm.nih.gov/pubmed/23140989 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC3558540 http://linkinghub.elsevier.com/retrieve/pii/S0168952512001692}

@@

\hypertarget{ref-Tillmann2017}{}
44. Tillmann T, Vaucher J, Okbay A, Pikhart H, Peasey A, Kubinova R, et
al. Education and coronary heart disease: a Mendelian randomization
study. bioRxiv {[}Internet{]}. 2017; Available from:
\url{http://www.biorxiv.org/content/early/2017/02/06/106237}

\hypertarget{ref-kotthoff2016auto}{}
45. Kotthoff L, Thornton C, Hoos HH, Hutter F, Leyton-Brown K. Auto-WEKA
2.0: Automatic model selection and hyperparameter optimization in WEKA.
Journal of Machine Learning Research. 2016;17:1--5.

\hypertarget{ref-Zhu2016}{}
46. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
Integration of summary data from GWAS and eQTL studies predicts complex
trait gene targets. Nature Genetics {[}Internet{]}. 2016
Mar;48(5):481--7. Available from:
\url{http://www.nature.com/doifinder/10.1038/ng.3538}

\hypertarget{ref-Richardson2017}{}
47. Richardson TG, Zheng J, Davey Smith G, Timpson NJ, Gaunt TR, Relton
CL, et al. Causal epigenome-wide association study identifies CpG sites
that influence cardiovascular disease risk. bioRxiv {[}Internet{]}.
2017; Available from:
\url{http://biorxiv.org/content/early/2017/04/29/132019}

\hypertarget{ref-Wang2017}{}
48. Wang L, Michoel T. Efficient And Accurate Causal Inference With
Hidden Confounders From Genome-Transcriptome Variation Data. bioRxiv
{[}Internet{]}. 2017; Available from:
\url{http://www.biorxiv.org/content/early/2017/04/19/128496}

\hypertarget{ref-Aalen2015}{}
49. Aalen OO, Valberg M, Grotmol T, Tretli S. Understanding variation in
disease risk: the elusive concept of frailty. International journal of
epidemiology {[}Internet{]}. 2015 Aug;44(4):1408--21. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/25501685\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4588855}{http://www.ncbi.nlm.nih.gov/pubmed/25501685 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4588855}

\hypertarget{ref-Noyce2017}{}
50. Noyce AJ, Kia DA, Hemani G, Nicolas A, Price TR, De Pablo-Fernandez
E, et al. Estimating the causal influence of body mass index on risk of
Parkinson disease: A Mendelian randomisation study. Brayne C, editor.
PLOS Medicine {[}Internet{]}. 2017 Jun;14(6):e1002314. Available from:
\url{http://dx.plos.org/10.1371/journal.pmed.1002314}

\hypertarget{ref-Burgess2014a}{}
51. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
multivariable Mendelian randomization to disentangle the causal effects
of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
Available from:
\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}

\hypertarget{ref-Juul2003}{}
52. Juul K, Tybjaerg-Hansen A, Marklund S, Heegaard NHH, Steffensen R,
Sillesen H, et al. Genetically Reduced Antioxidative Protection and
Increased Ischemic Heart Disease Risk: The Copenhagen City Heart Study.
Circulation {[}Internet{]}. 2003 Dec;109(1):59--65. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/14662715\%20http://circ.ahajournals.org/cgi/doi/10.1161/01.CIR.0000105720.28086.6C}{http://www.ncbi.nlm.nih.gov/pubmed/14662715 http://circ.ahajournals.org/cgi/doi/10.1161/01.CIR.0000105720.28086.6C}

\hypertarget{ref-VonHinke2016}{}
53. Hinke S von, Davey Smith G, Lawlor D, Propper C, Windmeijer F.
Genetic markers as instrumental variables. J Health Econ.
2016;45:131--48.

\hypertarget{ref-Staley2016}{}
54. Staley JR, Blackshaw J, Kamat MA, Ellis S, Surendran P, Sun BB, et
al. PhenoScanner: a database of human genotype--phenotype associations.
Bioinformatics {[}Internet{]}. 2016 Oct;32(20):3207--9. Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/27318201\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5048068\%20https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btw373}{http://www.ncbi.nlm.nih.gov/pubmed/27318201 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5048068 https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btw373}

\hypertarget{ref-Vanderweele2014}{}
55. Vanderweele TJ, Tchetgen Tchetgen EJ, Cornelis M, Kraft P.
Methodological challenges in mendelian randomization. Epidemiology
(Cambridge, Mass) {[}Internet{]}. 2014 May;25(3):427--35. Available
from: \url{http://www.ncbi.nlm.nih.gov/pubmed/24681576}
